---
title: "Generating allele frequencies of college squirrel populations"
---

```{r, message=FALSE}
sapply(c("ggplot2", "tidyr", "tibble", "dplyr", "purrr", "RColorBrewer"), require, character.only = TRUE)
source("utils.R") # plain R script containing the functions we wrote in the Wright-Fisher section
set.seed(23)
```

Here we will use our Wright-Fisher simulations to simulate the current allele frequencies among 5 different college campus squirrel populations: Boston University (B), UT-Austin (T), University of Michigan (M), Syracuse (S), and University of Florida (F). 

We will then filter out alleles that fall to a frequency of 0 in all populations. This will comprise our final dataset of alleles for the classroom exercise.

### Simulation Parameters:
```{r}
N <- 100
g <- 200
m <- 50
p_0 <- runif(n = m, min = 0, max = 0.25)
```

### Simulating each campus' squirrel allele frequencies:
```{r}
# Boston University
X_B <- wrightFisher(.N = N, .p = p_0, .g = g)
# UT-Austin
X_T <- wrightFisher(.N = N, .p = p_0, .g = g)
# University of Michigan
X_M <- wrightFisher(.N = N, .p = p_0, .g = g)
# Syracuse
X_S <- wrightFisher(.N = N, .p = p_0, .g = g)
# University of Florida
X_F <- wrightFisher(.N = N, .p = p_0, .g = g)
```
How many alleles are non-zero in at least one population?
```{r}
sum((X_B[, g] > 0) | (X_T[, g] > 0) | (X_M[, g] > 0) | 
      (X_S[, g] > 0) | (X_F[, g] > 0))
```
How many alleles are zero in all populations? (Should sum to $m$ with the previous set)
```{r}
zero_idxs <- which((X_B[, g] == 0) & (X_T[, g] == 0) & (X_M[, g] == 0) & 
                     (X_S[, g] == 0) & (X_F[, g] == 0))
length(zero_idxs)
```

Removing all-zero alleles:

```{r}
X_B <- X_B[-zero_idxs,]
X_T <- X_T[-zero_idxs,]
X_M <- X_M[-zero_idxs,]
X_S <- X_S[-zero_idxs,]
X_F <- X_F[-zero_idxs,]
```

Here are our final datasets for the classroom exercise, colored by frequency.

```{r}
nAlleles <- m - length(zero_idxs)
plotdf <- tibble(i = rep(seq(from = 10, to = 50, by = 10), each = nAlleles) + 5, 
                 j = rep(c(1:nAlleles), times = 5),
                 freq = c(X_B[, g], X_T[, g], X_M[, g], X_S[, g], X_F[, g]))
ggplot(plotdf, aes(x = i, y = j, fill = freq)) +
  geom_tile(width = 1) +
  geom_text(data = plotdf, aes(x = i + 3,
                               y = j,
                               label = round(freq, digits = 2))) +
  geom_text(data = plotdf, aes(x = i - 2,
                               y = j,
                               label = rev(j))) +
  scale_fill_distiller(palette = "Spectral") +
  theme_classic()
```

Saving these frequencies so they can be referenced permanently:

```{r}
# save(X_B, X_T, X_M, X_S, X_F, file = "Frequencies.RData")
```


